### 2020-09-30 submission/manuscript.tex
Update all references from Table S# Data Set S1, Sheet# in manuscript

We sequenced the V4 region of the 16S rRNA gene from fecal samples
collected from these mice after they acclimated to the University of
Michigan animal housing environment. We first examined the alpha
diversity across the 6 sources of mice. There was a significant
difference in the richness (i.e.~number of observed operational
taxonomic units (OTUs)), but not Shannon diversity index across the
sources of mice (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B and Tables
S1-2). Next, we compared the community structure of mice (Fig. 1C). The
source of mice and the interactions between the source and cage effects
explained most of the observed variation between fecal communities
(PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P} \textless{}
0.001; Fig. 1C and Table S3). Mice that are co-housed tend to have
similar gut microbiotas due to coprophagy (41). Since mice within the
same source were housed together, it was not surprising that the cage
effect also contributed to the observed community variation. There were
some differences between the 2 experiments we conducted, as the
experiment and cage effects significantly explained the observed
community variation for the Schloss and Young lab mouse colonies (Fig.
S1A-B and Table S4). However, most of the vendors also clustered by
experiment (Fig. S1C-D, F), suggesting there was some community
variation between the 2 experiments within each source, particularly for
Schloss, Young, and Envigo mice (Fig. S1G-H). After finding differences
at the community level, we next identified the bacteria that varied
between sources of mice. There were 268 OTUs with relative abundances
that were significantly different between the sources at baseline (Fig.
1D and Table S5). Though we saw differences between experiments at the
community level, there were no OTUs that were significantly different
between experiments within Schloss, Young, and Envigo mice at baseline
(all \emph{P} \textgreater{} 0.05). By using mice from six sources we
were able to increase the variation in the starting communities to
evaluate in a clindamycin-based CDI model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but clear the pathogen within 9
days, thus colonization is transient when treated with clindamycin alone
(21, 43). All mice were treated with 10 mg/kg clindamycin via
intraperitoneal injection and one day later challenged with
10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A). The day
after infection, \emph{C. difficile} was detectable in all mice at a
similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 2B). However, between 3 and
7 days post-infection, we observed variation in \emph{C. difficile}
levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
0.019; Fig. 2B and Table S6). This suggested the source of mice was
associated with \emph{C. difficile} clearance. While the colonization
dynamics were similar between the two experiments, the Schloss mice took
longer to clear \emph{C. difficile} in the first experiment compared to
the second and the Envigo mice took longer to clear \emph{C. difficile}
in the second experiment compared to the first (Fig. S2A-B). The change
in the mice's weight significantly varied across sources of mice with
the most weight loss occurring two days post-infection (Fig. 2C and
Table S7). There was also one Jackson and one Envigo mouse that died
between 1- and 3-days post-infection during the second experiment. Mice
obtained from Jackson, Taconic, and Envigo tended to lose more weight,
have higher \emph{C. difficile} CFU levels and take longer to clear the
infection compared to the other sources of mice (although there was
variation between experiments with Schloss and Envigo mice). This was
particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when
57\% of the mice were still colonized with \emph{C. difficile} (Fig.
S2E). By 9 days post-infection the majority of the mice from all sources
had cleared \emph{C. difficile} with the exception of 1 Taconic mouse
from the first experiment and 2 Envigo mice from the second experiment
(Fig. 2B). Thus, clindamycin rendered all mice susceptible to \emph{C.
difficile} 630 colonization, regardless of source, but there was
significant variation in disease phenotype across the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in fecal communities that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by community variation across the 6 sources
of mice. As expected, clindamycin treatment decreased the richness and
Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics between sources emerged
after clindamycin treatment, with Charles River mice having higher
richness and Shannon diversity than most of the other sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Tables
S1-2). The clindamycin treatment decreased the variation in community
structures between sources of mice. The source of mice and the
interactions between source and cage effects explained almost all of the
observed variation between communities (combined R\textsuperscript{2} =
0.99, \emph{P} \textless{} 0.001; Fig. 3C and Table S3). However, there
were only 18 OTUs with relative abundances that significantly varied
between sources after clindamycin treatment (Fig. 3D and Table S8).
Next, we identified the bacteria that shifted after clindamycin
treatment, regardless of source by analyzing paired fecal samples from
mice that were collected at baseline and after clindamycin treatment. We
identified 153 OTUs that were altered after clindamycin treatment in
most mice (Fig. 3E and Table S9). When we compared the list of
significant clindamycin impacted bacteria with the bacteria that varied
between sources post-clindamycin, we found 4 OTUs that were shared
between the lists (\emph{Enterobacteriaceae (OTU 1)},
\emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
\emph{Enterococcus} (OTU 23); Fig. 3D-E and Tables S8-9). Importantly,
some of the OTUs that varied between sources also shifted with
clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment (Fig. 3D), but only in Taconic mice.
\emph{Enterococcus} was primarily found in mice purchased from
commercial vendors and also increased in relative abundance after
clindamycin treatment (Fig. 3D). These findings demonstrate that
clindamycin had a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs continued to vary
between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse communities and
were also able to clear \emph{C. difficile} faster than the other
sources, diversity did not explain the observed variation in \emph{C.
difficile} colonization across sources. The Young and Schloss mice had
the lowest diversity 1 day post-infection and were able to clear
\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
source of mice and the interactions between source and cage effects
continued to explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
Table S3). One day after \emph{C. difficile} challenge, there were 44
OTUs with significantly different relative abundances across sources
(Fig. 4D and Table S10).

Throughout the experiment, the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Fecal samples from the same source
of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice had not recovered to their
baseline community structure (Fig. 4E). The distance to the baseline
community did not explain the variation in \emph{C. difficile} clearance
as the Schloss and Young mice had mostly cleared \emph{C. difficile},
but their communities were a greater distance from baseline 7 days
post-infection compared to the Jackson and Taconic mice that were still
colonized. In summary, mouse bacterial communities varied significantly
between sources throughout the course of the experiment and a consistent
subset of bacteria remained different between sources regardless of
clindamycin and \emph{C. difficile} challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we determined which taxa
were influencing the variation in \emph{C. difficile} colonization at
day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
regression models with either input bacterial community data from the 6
sources of mice at the baseline (day = -1), post-clindamycin (day = 0),
or post-infection (day = 1) timepoints of the experiment to predict
\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
models were better at predicting \emph{C. difficile} colonization status
on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
The model based on the post-clindamycin (AUROC = 0.78) community OTU
data performed significantly better than the baseline (AUROC = 0.72) or
the post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
\textless{} 0.001 for pairwise comparisons; Fig. S3C and Table S13).
Thus, we were able to use bacterial relative abundance data from the
time of \emph{C. difficile} challenge to differentiate mice that had
cleared \emph{C. difficile} before day 7 from the mice still colonized
with \emph{C. difficile} at that timepoint. This result suggests that
the bacterial community's response to clindamycin treatment had the
greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs were associated with \emph{C. difficile} colonization 7
days post-infection (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}; Fig. 5A). On day 0, the majority of these
OTUs were impacted by clindamycin and had relative abundances that were
close to the limit of detection (Fig. 5A). Next, we examined whether any
of the top 20 ranked OTUs from the post-clindamycin (day 0) model were
also important in the other 2 classification models based on baseline
(day -1) and 1 day post-infection community data. We identified 6 OTUs
that were important to the post-clindamycin model and either the
baseline or 1 day post-infection models (\emph{Enterobacteriaceae,
Ruminococcaceae, Lactobacillus, Bacteroides, Porphyromonadaceae,
Erysipelotrichaceae}; Table S14). Thus, a subset of bacterial OTUs were
important for determining \emph{C. difficile} colonization dynamics
across multiple timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we identified the OTUs from each model that varied between sources (Fig.
1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
important to the 3 models identified 14 OTUs associated with source, 21
OTUs associated with clindamycin treatment, and 6 OTUs associated with
both (Fig. 5B). Together, these results suggest that the initial
bacterial communities and their responses to clindamycin influenced the
clearance of \emph{C. difficile}.

Several OTUs that overlapped with our previous analyses appeared across
at least 2 models (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae}), so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Across the
9 days post-infection, there was at least 1 timepoint when the relative
abundances of these OTUs significantly varied between sources (Table
S15). Interestingly, there were no OTUs that emerged as consistently
enriched or depleted in mice that were colonized past 7 days
post-infection, suggesting that multiple bacteria influence \emph{C.
difficile} colonization dynamics.

@@

\textbf{Figure 1. Microbiota variation is high between mice from
different sources.} A-B. Number of observed OTUs (A) and Shannon
diversity index values (B) across sources of mice at baseline (day -1 of
the experiment). Differences between sources were analyzed by
Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
day of the experiment and the adjusted \emph{P} value was \textless{}
0.05 for panel A (Table S1). None of the \emph{P} values from pairwise
Wilcoxon comparisons between sources were significant after
Benjamini-Hochberg correction (Table S2). Gray lines represent the
median values for each source of mice. C. Principal Coordinates Analysis
(PCoA) of \(\theta_{YC}\) distances of baseline stool samples. Source
and the interaction between source and cage effects explained most of
the variation (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
\textless{} 0.001; Table S3). For A-C: each symbol represents the value
for a stool sample from an individual mouse, circles represent
experiment 1 mice and triangles represent experiment 2 mice. D. The
median (point) and interquantile range (colored lines) of the relative
abundances for the 20 most significant OTUs out of the 268 OTUs that
varied across sources at baseline by Kruskal-Wallis test with
Benjamini-Hochberg correction (Table S5).

@@

\textbf{Figure 2. Clindamycin is sufficient to promote \emph{C.
difficile} colonization in all mice, but clearance time varies across
sources.} A. Setup of the experimental timeline. Mice for the
experiments were obtained from 6 different sources: the Schloss (N = 8)
and Young lab (N = 9) colonies at the University of Michigan, the
Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
Biosciences (N = 8), and Envigo (N = 8). Mice that were ordered from
commercial vendors acclimated to the University of Michigan mouse
facility for 13 days prior to antibiotic administration. All mice were
administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before
challenge with \emph{C. difficile} 630 spores on day 0. Mice were
weighed and feces was collected daily through the end of the experiment
(9 days post-infection). Note: 3 mice died during course of experiment.
1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
stool measured over time (N = 20-49 mice per timepoint) via serial
dilutions. The black line represents the limit of detection for the
first serial dilution. CFU quantification data was not available for
each mouse due to early deaths, stool sampling difficulties, and not
plating all of the serial dilutions. C. Mouse weight change measured in
grams over time (N = 45-49 mice per timepoint), all mice were normalized
to the weight recorded 1 day before infection. For B-C: timepoints where
differences between sources of mice were statistically significant by
Kruskal-Wallis test with Benjamini-Hochberg correction for testing
across multiple days (Table S6 and Table S7) are reflected by the
asterisk above each timepoint (*, \emph{P} \textless{} 0.05). Lines
represent the median for each source and circles represent individual
mice from experiment 1 while triangles represent mice from experiment 2.

@@

\textbf{Figure 3. Clindamycin treatment alters bacteria in all sources,
but a subset of bacterial differences across sources persists.} A-B.
Number of observed OTUs (A) and Shannon diversity index values (B)
across sources of mice after clindamycin treatment (day 0). Differences
between sources were analyzed by Kruskal-Wallis test with
Benjamini-Hochberg correction for testing each day of the experiment and
the adjusted \emph{P} value was \textless{} 0.05 (Table S1). Significant
\emph{P} values from the pairwise Wilcoxon comparisons between sources
with Benjamini-Hochberg correction are displayed as the first initial of
each group compared to the group that they are listed above (Table S2).
C. PCoA of \(\theta_{YC}\) distances from stools collected
post-clindamycin. Source and the interaction between source and cage
effects explained most of the variation observed post-clindamycin
(PERMANOVA combined R\textsuperscript{2} = 0.99, \emph{P} \textless{}
0.001; Table S3). For A-C, each symbol represents a stool sample from an
individual mouse, with circles representing experiment 1 mice and
triangles representing experiment 2 mice. D. The median (point) and
interquantile range (colored lines) of the relative abundances for the
18 OTUs (Table S8) that varied between sources after clindamycin
treatment (day 0). E. The median (point) and interquantile range
(colored lines) of the top 10 most significant OTUs out of 153 with
relative abundances that changed because of the clindamycin treatment
(adjusted \emph{P} value \textless{} 0.05). Data were analyzed by paired
Wilcoxon signed rank test of mice that had paired sequence data for
baseline (day -1) and post-clindamycin (day 0) timepoints (N = 31), with
Benjamini-Hochberg correction for testing all identified OTUs (Table
S9). The gray vertical line indicates the limit of detection.

@@

\textbf{Figure 5. Bacteria that influenced whether mice cleared \emph{C.
difficile} by day 7.} A. Post-clindamycin (day 0) relative abundance
data for the 10 OTUs with the highest rankings based on feature weights
in the post-clindamycin (day 0) classification model. Red font
represents OTUs that correlated with \emph{C. difficile} colonization
and blue font represents OTUs that correlated with clearance. Symbols
represent the relative abundance data for an individual mouse. Gray bars
indicate the median relative abundances for each source. The gray
horizontal lines indicate the limit of detection. B. Venn diagram that
combines OTUs that were important to the day -1, 0, and 1 classification
models ( Fig. S4, Table S14) and either overlapped with taxa that varied
across sources at the same timepoint, were impacted by clindamycin
treatment, or both. Bold OTUs were important to more than 1
classification model.

>>>>

We sequenced the V4 region of the 16S rRNA gene from fecal samples
collected from these mice after they acclimated to the University of
Michigan animal housing environment. We first examined the alpha
diversity across the 6 sources of mice. There was a significant
difference in the richness (i.e.~number of observed operational
taxonomic units (OTUs)), but not Shannon diversity index across the
sources of mice (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B and Data
Set S1, Sheets 1-2). Next, we compared the community structure of mice
(Fig. 1C). The source of mice and the interactions between the source
and cage effects explained most of the observed variation between fecal
communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
\textless{} 0.001; Fig. 1C and Data Set S1, Sheet 3). Mice that are
co-housed tend to have similar gut microbiotas due to coprophagy (41).
Since mice within the same source were housed together, it was not
surprising that the cage effect also contributed to the observed
community variation. There were some differences between the 2
experiments we conducted, as the experiment and cage effects
significantly explained the observed community variation for the Schloss
and Young lab mouse colonies (Fig. S1A-B and Data Set S1, Sheet 4).
However, most of the vendors also clustered by experiment (Fig. S1C-D,
F), suggesting there was some community variation between the 2
experiments within each source, particularly for Schloss, Young, and
Envigo mice (Fig. S1G-H). After finding differences at the community
level, we next identified the bacteria that varied between sources of
mice. There were 268 OTUs with relative abundances that were
significantly different between the sources at baseline (Fig. 1D and
Data Set S1, Sheet 5). Though we saw differences between experiments at
the community level, there were no OTUs that were significantly
different between experiments within Schloss, Young, and Envigo mice at
baseline (all \emph{P} \textgreater{} 0.05). By using mice from six
sources we were able to increase the variation in the starting
communities to evaluate in a clindamycin-based CDI model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but clear the pathogen within 9
days, thus colonization is transient when treated with clindamycin alone
(21, 43). All mice were treated with 10 mg/kg clindamycin via
intraperitoneal injection and one day later challenged with
10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A). The day
after infection, \emph{C. difficile} was detectable in all mice at a
similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 2B). However, between 3 and
7 days post-infection, we observed variation in \emph{C. difficile}
levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
0.019; Fig. 2B and Data Set S1, Sheet 6). This suggested the source of
mice was associated with \emph{C. difficile} clearance. While the
colonization dynamics were similar between the two experiments, the
Schloss mice took longer to clear \emph{C. difficile} in the first
experiment compared to the second and the Envigo mice took longer to
clear \emph{C. difficile} in the second experiment compared to the first
(Fig. S2A-B). The change in the mice's weight significantly varied
across sources of mice with the most weight loss occurring two days
post-infection (Fig. 2C and Data Set S1, Sheet 7). There was also one
Jackson and one Envigo mouse that died between 1- and 3-days
post-infection during the second experiment. Mice obtained from Jackson,
Taconic, and Envigo tended to lose more weight, have higher \emph{C.
difficile} CFU levels and take longer to clear the infection compared to
the other sources of mice (although there was variation between
experiments with Schloss and Envigo mice). This was particularly evident
7 days post-infection (Fig. 2B-C, Fig. S2C-D), when 57\% of the mice
were still colonized with \emph{C. difficile} (Fig. S2E). By 9 days
post-infection the majority of the mice from all sources had cleared
\emph{C. difficile} with the exception of 1 Taconic mouse from the first
experiment and 2 Envigo mice from the second experiment (Fig. 2B). Thus,
clindamycin rendered all mice susceptible to \emph{C. difficile} 630
colonization, regardless of source, but there was significant variation
in disease phenotype across the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in fecal communities that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by community variation across the 6 sources
of mice. As expected, clindamycin treatment decreased the richness and
Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics between sources emerged
after clindamycin treatment, with Charles River mice having higher
richness and Shannon diversity than most of the other sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Data Set S1,
Sheets 1-2). The clindamycin treatment decreased the variation in
community structures between sources of mice. The source of mice and the
interactions between source and cage effects explained almost all of the
observed variation between communities (combined R\textsuperscript{2} =
0.99, \emph{P} \textless{} 0.001; Fig. 3C and Data Set S1, Sheet 3).
However, there were only 18 OTUs with relative abundances that
significantly varied between sources after clindamycin treatment (Fig.
3D and Data Set S1, Sheet 8). Next, we identified the bacteria that
shifted after clindamycin treatment, regardless of source by analyzing
paired fecal samples from mice that were collected at baseline and after
clindamycin treatment. We identified 153 OTUs that were altered after
clindamycin treatment in most mice (Fig. 3E and Data Set S1, Sheet 9).
When we compared the list of significant clindamycin impacted bacteria
with the bacteria that varied between sources post-clindamycin, we found
4 OTUs that were shared between the lists (\emph{Enterobacteriaceae}
(OTU 1), \emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
\emph{Enterococcus} (OTU 23); Fig. 3D-E and Data Set S1, Sheets 8-9).
Importantly, some of the OTUs that varied between sources also shifted
with clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment (Fig. 3D), but only in Taconic mice.
\emph{Enterococcus} was primarily found in mice purchased from
commercial vendors and also increased in relative abundance after
clindamycin treatment (Fig. 3D). These findings demonstrate that
clindamycin had a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs continued to vary
between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Data Set S1,
Sheets 1-2). Although the Charles River mice had more diverse
communities and were also able to clear \emph{C. difficile} faster than
the other sources, diversity did not explain the observed variation in
\emph{C. difficile} colonization across sources. The Young and Schloss
mice had the lowest diversity 1 day post-infection and were able to
clear \emph{C. difficile} earlier than Jackson, Taconic and Envigo mice.
The source of mice and the interactions between source and cage effects
continued to explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
Data Set S1, Sheet 3). One day after \emph{C. difficile} challenge,
there were 44 OTUs with significantly different relative abundances
across sources (Fig. 4D and Data Set S1, Sheet 10).

Throughout the experiment, the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Data Set S1, Sheet 11). Fecal samples from the
same source of mice continued to cluster closely to each other
throughout the experiment. By 7 days post-infection, when approximately
43\% mice had cleared \emph{C. difficile}, most of the mice had not
recovered to their baseline community structure (Fig. 4E). The distance
to the baseline community did not explain the variation in \emph{C.
difficile} clearance as the Schloss and Young mice had mostly cleared
\emph{C. difficile}, but their communities were a greater distance from
baseline 7 days post-infection compared to the Jackson and Taconic mice
that were still colonized. In summary, mouse bacterial communities
varied significantly between sources throughout the course of the
experiment and a consistent subset of bacteria remained different
between sources regardless of clindamycin and \emph{C. difficile}
challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we determined which taxa
were influencing the variation in \emph{C. difficile} colonization at
day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
regression models with either input bacterial community data from the 6
sources of mice at the baseline (day = -1), post-clindamycin (day = 0),
or post-infection (day = 1) timepoints of the experiment to predict
\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
models were better at predicting \emph{C. difficile} colonization status
on day 7 than random chance (all \emph{P} \textless{} 0.001, Data Set
S1, Sheet 12). The model based on the post-clindamycin (AUROC = 0.78)
community OTU data performed significantly better than the baseline
(AUROC = 0.72) or the post-infection (AUROC = 0.67) models
(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons;
Fig. S3C and Data Set S1, Sheet 13). Thus, we were able to use bacterial
relative abundance data from the time of \emph{C. difficile} challenge
to differentiate mice that had cleared \emph{C. difficile} before day 7
from the mice still colonized with \emph{C. difficile} at that
timepoint. This result suggests that the bacterial community's response
to clindamycin treatment had the greatest influence on subsequent
\emph{C. difficile} colonization dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Data Set S1, Sheet 14). First, we looked at OTUs
from the model with the best performance, which was based on the
post-clindamycin treatment (day 0) bacterial community data. Out of the
10 highest ranked OTUs, 7 OTUs were associated with \emph{C. difficile}
colonization 7 days post-infection (\emph{Bacteroides},
\emph{Escherichia/Shigella}, 2 \emph{Lachnospiraceae},
\emph{Lactobacillus}, \emph{Porphyromonadaceae}, and
\emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}; Fig. 5A). On day 0, the majority of these
OTUs were impacted by clindamycin and had relative abundances that were
close to the limit of detection (Fig. 5A). Next, we examined whether any
of the top 20 ranked OTUs from the post-clindamycin (day 0) model were
also important in the other 2 classification models based on baseline
(day -1) and 1 day post-infection community data. We identified 6 OTUs
that were important to the post-clindamycin model and either the
baseline or 1 day post-infection models (\emph{Enterobacteriaceae,
Ruminococcaceae, Lactobacillus, Bacteroides, Porphyromonadaceae,
Erysipelotrichaceae}; Data Set S1, Sheet 14). Thus, a subset of
bacterial OTUs were important for determining \emph{C. difficile}
colonization dynamics across multiple timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we identified the OTUs from each model that varied between sources (Fig.
1D, 3D, 4D and Data Set S1, Sheets 5, 8, and 10) or were impacted by
clindamycin treatment (Fig. 3E and Data Set S1, Sheet 9; Fig. S4).
Comparing the features important to the 3 models identified 14 OTUs
associated with source, 21 OTUs associated with clindamycin treatment,
and 6 OTUs associated with both (Fig. 5B). Together, these results
suggest that the initial bacterial communities and their responses to
clindamycin influenced the clearance of \emph{C. difficile}.

Several OTUs that overlapped with our previous analyses appeared across
at least 2 models (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae}), so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Across the
9 days post-infection, there was at least 1 timepoint when the relative
abundances of these OTUs significantly varied between sources (Data Set
S1, Sheet 15). Interestingly, there were no OTUs that emerged as
consistently enriched or depleted in mice that were colonized past 7
days post-infection, suggesting that multiple bacteria influence
\emph{C. difficile} colonization dynamics.

@@

\textbf{Figure 1. Microbiota variation is high between mice from
different sources.} A-B. Number of observed OTUs (A) and Shannon
diversity index values (B) across sources of mice at baseline (day -1 of
the experiment). Differences between sources were analyzed by
Kruskal-Wallis test with Benjamini-Hochberg correction for testing each
day of the experiment and the adjusted \emph{P} value was \textless{}
0.05 for panel A (Data Set S1, Sheet 1). None of the \emph{P} values
from pairwise Wilcoxon comparisons between sources were significant
after Benjamini-Hochberg correction (Data Set S1, Sheet 2). Gray lines
represent the median values for each source of mice. C. Principal
Coordinates Analysis (PCoA) of \(\theta_{YC}\) distances of baseline
stool samples. Source and the interaction between source and cage
effects explained most of the variation (PERMANOVA combined
R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001; Data Set S1,
Sheet 3). For A-C: each symbol represents the value for a stool sample
from an individual mouse, circles represent experiment 1 mice and
triangles represent experiment 2 mice. D. The median (point) and
interquantile range (colored lines) of the relative abundances for the
20 most significant OTUs out of the 268 OTUs that varied across sources
at baseline by Kruskal-Wallis test with Benjamini-Hochberg correction
(Data Set S1, Sheet 5).

@@

\textbf{Figure 2. Clindamycin is sufficient to promote \emph{C.
difficile} colonization in all mice, but clearance time varies across
sources.} A. Setup of the experimental timeline. Mice for the
experiments were obtained from 6 different sources: the Schloss (N = 8)
and Young lab (N = 9) colonies at the University of Michigan, the
Jackson Laboratory (N = 8), Charles River Laboratory (N = 8), Taconic
Biosciences (N = 8), and Envigo (N = 8). Mice that were ordered from
commercial vendors acclimated to the University of Michigan mouse
facility for 13 days prior to antibiotic administration. All mice were
administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before
challenge with \emph{C. difficile} 630 spores on day 0. Mice were
weighed and feces was collected daily through the end of the experiment
(9 days post-infection). Note: 3 mice died during course of experiment.
1 Taconic mouse prior to infection and 1 Jackson and 1 Envigo mouse
between 1- and 3-days post-infection. B. \emph{C. difficile} CFU/gram
stool measured over time (N = 20-49 mice per timepoint) via serial
dilutions. The black line represents the limit of detection for the
first serial dilution. CFU quantification data was not available for
each mouse due to early deaths, stool sampling difficulties, and not
plating all of the serial dilutions. C. Mouse weight change measured in
grams over time (N = 45-49 mice per timepoint), all mice were normalized
to the weight recorded 1 day before infection. For B-C: timepoints where
differences between sources of mice were statistically significant by
Kruskal-Wallis test with Benjamini-Hochberg correction for testing
across multiple days (Data Set S1, Sheets 6-7) are reflected by the
asterisk above each timepoint (*, \emph{P} \textless{} 0.05). Lines
represent the median for each source and circles represent individual
mice from experiment 1 while triangles represent mice from experiment 2.

@@

\textbf{Figure 3. Clindamycin treatment alters bacteria in all sources,
but a subset of bacterial differences across sources persists.} A-B.
Number of observed OTUs (A) and Shannon diversity index values (B)
across sources of mice after clindamycin treatment (day 0). Differences
between sources were analyzed by Kruskal-Wallis test with
Benjamini-Hochberg correction for testing each day of the experiment and
the adjusted \emph{P} value was \textless{} 0.05 (Data Set S1, Sheet 1).
Significant \emph{P} values from the pairwise Wilcoxon comparisons
between sources with Benjamini-Hochberg correction are displayed as the
first initial of each group compared to the group that they are listed
above (Data Set S1, Sheet 2). C. PCoA of \(\theta_{YC}\) distances from
stools collected post-clindamycin. Source and the interaction between
source and cage effects explained most of the variation observed
post-clindamycin (PERMANOVA combined R\textsuperscript{2} = 0.99,
\emph{P} \textless{} 0.001; Data Set S1, Sheet 3). For A-C, each symbol
represents a stool sample from an individual mouse, with circles
representing experiment 1 mice and triangles representing experiment 2
mice. D. The median (point) and interquantile range (colored lines) of
the relative abundances for the 18 OTUs (Data Set S1, Sheet 8) that
varied between sources after clindamycin treatment (day 0). E. The
median (point) and interquantile range (colored lines) of the top 10
most significant OTUs out of 153 with relative abundances that changed
because of the clindamycin treatment (adjusted \emph{P} value
\textless{} 0.05). Data were analyzed by paired Wilcoxon signed rank
test of mice that had paired sequence data for baseline (day -1) and
post-clindamycin (day 0) timepoints (N = 31), with Benjamini-Hochberg
correction for testing all identified OTUs (Data Set S1, Sheet 9). The
gray vertical line indicates the limit of detection.

@@

\textbf{Figure 5. Bacteria that influenced whether mice cleared \emph{C.
difficile} by day 7.} A. Post-clindamycin (day 0) relative abundance
data for the 10 OTUs with the highest rankings based on feature weights
in the post-clindamycin (day 0) classification model. Red font
represents OTUs that correlated with \emph{C. difficile} colonization
and blue font represents OTUs that correlated with clearance. Symbols
represent the relative abundance data for an individual mouse. Gray bars
indicate the median relative abundances for each source. The gray
horizontal lines indicate the limit of detection. B. Venn diagram that
combines OTUs that were important to the day -1, 0, and 1 classification
models ( Fig. S4, Data Set S1, Sheet 14) and either overlapped with taxa
that varied across sources at the same timepoint, were impacted by
clindamycin treatment, or both. Bold OTUs were important to more than 1
classification model.
